A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-01 paper.Rmd
developing ...

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("region", "kon", "birthdate", "deathdate", "exit", "event")]
x <- head(infdat, 5)
tbl(x, caption = "A few infants in the data set.")
```

@@

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}

tid <- toTpch(Surv(enter, exit, event) ~ week, data =  infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=5,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("subreg", "sex", "birthdate", "exit", "event")]
x <- head(infdat, 5)
tbl(x, caption = "A few infants in the data set.")
```

@@

The BP procedure consists of two steps, to begin with:

1.   Apply the log-cube transformation to the age data, and

2.   plot the cumulative sum of the number of deaths vs. age.

We plot without the first step in the left hand panel and with it in the
right hand panel, see Figure \@ref(fig:ggplot2).

```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 7}
nbths <- NROW(infdat)
oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     main = "Untransformed", ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
plot(g(xt), ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = g(att), labels = att)
box()
##
par(oldpar)
```

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```
